---
title: "Course setup"
format: html
---

This course contains a few hands-on activities to aid understanding for some concepts we will cover. These have been designed to be carried out in R. To follow along, please make sure you have the following ready before coming to class.

::: callout-important
In case you are unable to download the specified packages or dataset, please reach out to the instructors **before** class.
:::

## 1. Environment

1.  Ensure you have the latest compatible versions of RStudio and R installed for your computer. If you have not previously used R/RStudio, you can install them using the Self-Service desktop application or by reaching out directly to JAX IT.

2.  Make sure you have the following packages installed. Simply copy the code chunk and run it in the RStudio console:

```{r}
#| eval: false

install.packages(c(
  "tidyverse",
  "factoextra",
  "Rtsne",
  "uwot",
  "patchwork",
  "Seurat"
))
```

If `Seurat` doesn't install correctly, you can install it through BioConductor as follows:

```{r}
#| eval: false

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("Seurat")
```

That should do the trick!

3.  To make sure the packages installed correctly, go ahead and try loading them in by running the following in the console:

```{r}
#| eval: false

library(tidyverse)
library(factoextra)
library(Rtsne)
library(uwot)
library(patchwork)
library(Seurat)
```

## 2. Data

We will be using the same dataset throughout the course. The original study sequenced 2,700 peripheral blood mononuclear cells (PBMCs) from a healthy donor, which is freely available as the [PBMC3k dataset from 10X Genomics](https://www.10xgenomics.com/datasets/3-k-pbm-cs-from-a-healthy-donor-1-standard-1-1-0).

A light-weight subset of this dataset is available as `pbmc_small`, which contains expression data for 200+ genes measured for 80 cells. utilizing a subset of the data helps ensure that all code can be easily and quickly run on most participants' systems locally.

You do not need to download this dataset ahead of class, as it can be accessed after loading the `Seurat` package in our session.

## 3. Code

A Quarto document has been prepared and is available for participants to follow along with the workshop. It is recommended to download and work on this [participant notebook](https://raw.githubusercontent.com/sonalkumarr/linear-algebra-for-biologists/main/scripts/linear_algebra_participant.qmd){download="linear_algebra_participant.qmd"}. You will be using it throughout the workshop, and you can run through it line by line with the instructor and make any notes as needed.

This is all that you need to get started!

::: callout-note
If you'd like to follow along with the class in real time, then please plan to come with these installed/downloaded. As this is a short workshop, we will not have time to troubleshoot software installations!
:::
