---
title: "Linear Algebra for Biologists"
subtitle: "Instructor Workshop Notebook"
author: "Sonal Kumar"
format:
  html:
    toc: true
    code-fold: false
execute:
  eval: false
  echo: true
  warning: false
  message: false
---

# Instructor notes

This notebook is designed to accompany the workshop content.  
Do **not** spend the full workshop walking through every line of code.

Suggested use:

- Run selected chunks live.
- Use plots as teaching anchors.
- Pause often for interpretation.
- Skip optional chunks if running behind.

Main teaching story:

> Biological data can be represented as a matrix.  
> Once data is a matrix, we can compare samples, transform the data, reduce dimensionality, and visualize structure.

Dataset: PBMC3k single-cell RNA-seq data.

We will treat:

- cells as observations/samples
- genes as features/variables
- expression values as measurements

# Setup

```{r}
library(tidyverse)
library(Seurat)
library(patchwork)
library(uwot)
library(Rtsne)

set.seed(123)
```

```{r}
theme_set(theme_minimal(base_size = 12))
```

# Module 1: Understanding biological data structure

## Goal

By the end of this module, participants should understand:

- biological data as a matrix
- rows and columns as observations and features
- cells as vectors
- similarity and distance between cells

## Load dataset

```{r}
options(timeout = 300) # Increases timeout to 5 minutes
SeuratData::InstallData("pbmc3k")


SeuratData::InstallData("pbmc3k")
pbmc <- SeuratData::LoadData("pbmc3k")
```


Instructor note:  
Use a preprocessed PBMC3k object for the workshop. Avoid making participants run the full Seurat preprocessing workflow unless this is explicitly part of the course.

```{r}
pbmc <- readRDS("data/pbmc3k_clean.rds")
pbmc
```

## Inspect metadata

```{r}
head(pbmc@meta.data)
```

```{r}
table(pbmc$seurat_annotations)
```

Instructor pause:

Ask:

- What does one row in the metadata represent?
- What kinds of information do we have about each cell?
- Which variables are biological labels?
- Which variables might be technical?

## Extract expression matrix

```{r}
expr <- GetAssayData(pbmc, assay = "RNA", layer = "data")
expr[1:5, 1:5]
```

```{r}
dim(expr)
```

Instructor note:

Seurat stores this matrix as:

```text
genes × cells
```

For most of our linear algebra intuition, it is easier to think in terms of:

```text
cells × genes
```

So we transpose the matrix.

```{r}
expr_t <- t(as.matrix(expr))
dim(expr_t)
```

Instructor pause:

Ask:

- What does one row represent now?
- What does one column represent now?
- What is one cell represented as mathematically?

Answer:

Each cell is a vector of gene expression values.

## Select variable genes

Instructor note:

We do not need every gene for teaching. Using all genes slows things down and adds noise.

```{r}
variable_genes <- VariableFeaturePlot(pbmc)[1:1000]
expr_var <- expr_t[, variable_genes]

dim(expr_var)
```

## Visualize a small part of the matrix

```{r}
expr_small <- expr_var[1:10, 1:10]
expr_small
```

Instructor pause:

Ask:

- Is this a small matrix or a slice of a larger matrix?
- What would a row vector mean?
- What would a column vector mean?

## Look at expression distribution

```{r}
expr_long_sample <- expr_var[
  sample(rownames(expr_var), 300),
  sample(colnames(expr_var), 50)
] %>%
  as.data.frame() %>%
  rownames_to_column("cell") %>%
  pivot_longer(-cell, names_to = "gene", values_to = "expression")

ggplot(expr_long_sample, aes(x = expression)) +
  geom_histogram(bins = 50) +
  labs(
    title = "Distribution of expression values",
    x = "Log-normalized expression",
    y = "Count"
  )
```

Instructor pause:

Ask:

- Are most values high or low?
- What does this suggest about biological data?
- Is this matrix dense or sparse-ish?

# Similarity and distance

## Subset cells for distance calculations

Instructor note:

Pairwise distances across all cells can be slow and visually overwhelming. Use a subset.

```{r}
cell_subset <- sample(rownames(expr_var), 250)
expr_subset <- expr_var[cell_subset, ]
```

## Compute distances between cells

```{r}
cell_dist <- dist(expr_subset, method = "euclidean")
cell_dist_matrix <- as.matrix(cell_dist)

cell_dist_matrix[1:5, 1:5]
```

Instructor explanation:

Each value represents how far apart two cells are based on their gene expression vectors.

Instructor pause:

Ask:

- What does a distance of 0 mean?
- Why is the diagonal 0?
- Why is the matrix symmetric?

## Hierarchical clustering

```{r}
hc <- hclust(cell_dist)

plot(hc, labels = FALSE, main = "Hierarchical clustering of cells")
```

Instructor pause:

Ask:

- What is being clustered?
- What information was used to cluster cells?
- Do we know yet whether these clusters are biologically meaningful?

## Heatmap of cell distances

```{r}
heatmap(
  cell_dist_matrix,
  Rowv = as.dendrogram(hc),
  Colv = as.dendrogram(hc),
  scale = "none",
  labRow = FALSE,
  labCol = FALSE,
  main = "Cell-cell distance matrix"
)
```

Instructor note:

This is a good moment to connect:

- similarity
- distance
- clustering
- visual structure

# Module 2: Transformations

## Goal

By the end of this module, participants should understand:

- centering
- scaling
- why preprocessing changes data geometry
- why transformations affect downstream analyses

## Gene means and variances

```{r}
gene_summary <- tibble(
  gene = colnames(expr_var),
  mean_expression = apply(expr_var, 2, mean),
  variance = apply(expr_var, 2, var),
  sd = apply(expr_var, 2, sd)
)

head(gene_summary)
```

```{r}
ggplot(gene_summary, aes(x = mean_expression, y = variance)) +
  geom_point(alpha = 0.5, size = 1) +
  labs(
    title = "Mean expression vs variance",
    x = "Mean expression",
    y = "Variance"
  )
```

Instructor pause:

Ask:

- Do all genes vary equally?
- Why might highly variable genes dominate analyses?
- Is high variance always biological?

## Center data

```{r}
expr_centered <- scale(expr_var, center = TRUE, scale = FALSE)
```

Check one gene before and after centering:

```{r}
example_gene <- colnames(expr_var)[1]

mean(expr_var[, example_gene])
mean(expr_centered[, example_gene])
```

Instructor explanation:

Centering shifts each gene so its mean is 0.

## Scale data

```{r}
expr_scaled <- scale(expr_var, center = TRUE, scale = TRUE)
```

Check one gene before and after scaling:

```{r}
sd(expr_var[, example_gene])
sd(expr_scaled[, example_gene])
```

Instructor explanation:

Scaling makes each gene have standard deviation 1.

## Compare raw, centered, and scaled values

```{r}
compare_gene_df <- tibble(
  cell = rownames(expr_var),
  raw = expr_var[, example_gene],
  centered = expr_centered[, example_gene],
  scaled = expr_scaled[, example_gene]
) %>%
  pivot_longer(-cell, names_to = "version", values_to = "value")

ggplot(compare_gene_df, aes(x = value)) +
  geom_histogram(bins = 40) +
  facet_wrap(~ version, scales = "free") +
  labs(
    title = paste("Effect of centering and scaling on", example_gene),
    x = "Value",
    y = "Count"
  )
```

Instructor pause:

Ask:

- What changed after centering?
- What changed after scaling?
- Did the relative ordering of cells change for this gene?

# Module 3: Dimensionality reduction with PCA

## Goal

By the end of this module, participants should understand:

- why high-dimensional data is hard to visualize
- PCA as a rotation of the data
- PCs as directions of variation
- variance explained
- loadings as gene contributions

## Run PCA on unscaled data

```{r}
pca_unscaled <- prcomp(expr_var, center = TRUE, scale. = FALSE)
```

```{r}
pca_unscaled_df <- as_tibble(pca_unscaled$x[, 1:2], rownames = "cell") %>%
  left_join(
    pbmc@meta.data %>% rownames_to_column("cell"),
    by = "cell"
  )
```

```{r}
ggplot(pca_unscaled_df, aes(x = PC1, y = PC2, color = seurat_annotations)) +
  geom_point(size = 0.7, alpha = 0.7) +
  labs(
    title = "PCA on centered, unscaled data",
    color = "Cell type"
  )
```

Instructor pause:

Ask:

- What structure do we see?
- Are cell types separated?
- What might dominate this PCA?

## Run PCA on scaled data

```{r}
pca_scaled <- prcomp(expr_scaled, center = FALSE, scale. = FALSE)
```

```{r}
pca_scaled_df <- as_tibble(pca_scaled$x[, 1:2], rownames = "cell") %>%
  left_join(
    pbmc@meta.data %>% rownames_to_column("cell"),
    by = "cell"
  )
```

```{r}
ggplot(pca_scaled_df, aes(x = PC1, y = PC2, color = seurat_annotations)) +
  geom_point(size = 0.7, alpha = 0.7) +
  labs(
    title = "PCA on scaled data",
    color = "Cell type"
  )
```

Instructor pause:

Ask:

- How did scaling change the PCA?
- Which version is easier to interpret?
- Does scaling always make things better?

## Compare PCA with and without labels

```{r}
ggplot(pca_scaled_df, aes(x = PC1, y = PC2)) +
  geom_point(size = 0.7, alpha = 0.7) +
  labs(title = "PCA without cell-type labels")
```

```{r}
ggplot(pca_scaled_df, aes(x = PC1, y = PC2, color = seurat_annotations)) +
  geom_point(size = 0.7, alpha = 0.7) +
  labs(
    title = "PCA with cell-type labels",
    color = "Cell type"
  )
```

Instructor note:

This is a good place to emphasize:

> Dimensionality reduction does not know the labels.  
> We add labels afterward to interpret the structure.

## Variance explained

```{r}
variance_explained <- pca_scaled$sdev^2 / sum(pca_scaled$sdev^2)

variance_df <- tibble(
  PC = seq_along(variance_explained),
  variance_explained = variance_explained
)

ggplot(variance_df[1:30, ], aes(x = PC, y = variance_explained)) +
  geom_col() +
  labs(
    title = "Variance explained by principal components",
    x = "Principal component",
    y = "Proportion of variance explained"
  )
```

Instructor pause:

Ask:

- Why does PC1 explain the most variance?
- Why do later PCs explain less?
- How many PCs would you keep?

## Cumulative variance explained

```{r}
variance_df <- variance_df %>%
  mutate(cumulative_variance = cumsum(variance_explained))

ggplot(variance_df[1:50, ], aes(x = PC, y = cumulative_variance)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Cumulative variance explained",
    x = "Number of PCs",
    y = "Cumulative variance explained"
  )
```

Instructor note:

In single-cell data, variance is often spread across many PCs.  
This is normal and useful to discuss.

## PCA loadings

```{r}
loadings_df <- tibble(
  gene = rownames(pca_scaled$rotation),
  PC1 = pca_scaled$rotation[, 1],
  PC2 = pca_scaled$rotation[, 2]
)

loadings_df %>%
  arrange(desc(abs(PC1))) %>%
  slice_head(n = 10)
```

```{r}
loadings_df %>%
  arrange(desc(abs(PC2))) %>%
  slice_head(n = 10)
```

Instructor explanation:

Loadings tell us which genes contribute strongly to a principal component.

Instructor pause:

Ask:

- If a gene has a large loading on PC1, what does that mean?
- Does PCA know which genes are biologically important?
- How would we interpret these genes carefully?

# Module 4: Beyond PCA — UMAP and t-SNE

## Goal

By the end of this module, participants should understand:

- PCA, UMAP, and t-SNE are all dimensionality reduction methods
- PCA emphasizes global linear structure
- UMAP and t-SNE emphasize local neighborhood structure
- visualizations are tools, not proof

## Use PCs as input

Instructor note:

For UMAP and t-SNE, we usually do not use all genes directly.  
We use top PCs as a lower-dimensional representation of the data.

```{r}
pca_coords <- pca_scaled$x[, 1:20]
dim(pca_coords)
```

## Run UMAP

```{r}
umap_coords <- umap(
  pca_coords,
  n_neighbors = 15,
  min_dist = 0.3,
  metric = "euclidean"
)
```

```{r}
umap_df <- as_tibble(umap_coords, .name_repair = "unique") %>%
  rename(UMAP1 = ...1, UMAP2 = ...2) %>%
  mutate(cell = rownames(pca_coords)) %>%
  left_join(
    pbmc@meta.data %>% rownames_to_column("cell"),
    by = "cell"
  )
```

```{r}
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = seurat_annotations)) +
  geom_point(size = 0.7, alpha = 0.7) +
  labs(
    title = "UMAP of PBMC3k",
    color = "Cell type"
  )
```

Instructor pause:

Ask:

- What looks different from PCA?
- Are clusters more separated?
- Does more separation mean more truth?

## Run t-SNE

Instructor note:

This may take a little longer. Pre-run if needed.

```{r}
tsne_coords <- Rtsne(
  pca_coords,
  dims = 2,
  perplexity = 30,
  verbose = FALSE
)$Y
```

```{r}
tsne_df <- as_tibble(tsne_coords, .name_repair = "unique") %>%
  rename(tSNE1 = ...1, tSNE2 = ...2) %>%
  mutate(cell = rownames(pca_coords)) %>%
  left_join(
    pbmc@meta.data %>% rownames_to_column("cell"),
    by = "cell"
  )
```

```{r}
ggplot(tsne_df, aes(x = tSNE1, y = tSNE2, color = seurat_annotations)) +
  geom_point(size = 0.7, alpha = 0.7) +
  labs(
    title = "t-SNE of PBMC3k",
    color = "Cell type"
  )
```

## Compare PCA, UMAP, and t-SNE

```{r}
p_pca <- ggplot(pca_scaled_df, aes(x = PC1, y = PC2, color = seurat_annotations)) +
  geom_point(size = 0.5, alpha = 0.7) +
  labs(title = "PCA", color = "Cell type") +
  theme(legend.position = "none")

p_umap <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = seurat_annotations)) +
  geom_point(size = 0.5, alpha = 0.7) +
  labs(title = "UMAP", color = "Cell type") +
  theme(legend.position = "none")

p_tsne <- ggplot(tsne_df, aes(x = tSNE1, y = tSNE2, color = seurat_annotations)) +
  geom_point(size = 0.5, alpha = 0.7) +
  labs(title = "t-SNE", color = "Cell type")

p_pca + p_umap + p_tsne
```

Instructor pause:

Ask:

- Which method separates clusters most clearly?
- Which method is easiest to overinterpret?
- Which method preserves global structure best?
- Are distances between clusters meaningful in all plots?

# Module 5: How preprocessing changes interpretation

## Goal

By the end of this module, participants should understand:

- analysis choices affect visual output
- different parameter choices can change interpretation
- plots should be interpreted critically

## Try fewer variable genes

```{r}
variable_genes_200 <- VariableFeatures(pbmc)[1:200]
expr_200 <- expr_t[, variable_genes_200]
expr_200_scaled <- scale(expr_200, center = TRUE, scale = TRUE)

pca_200 <- prcomp(expr_200_scaled, center = FALSE, scale. = FALSE)
pca_200_df <- as_tibble(pca_200$x[, 1:2], rownames = "cell") %>%
  left_join(
    pbmc@meta.data %>% rownames_to_column("cell"),
    by = "cell"
  )
```

```{r}
ggplot(pca_200_df, aes(x = PC1, y = PC2, color = seurat_annotations)) +
  geom_point(size = 0.7, alpha = 0.7) +
  labs(
    title = "PCA using top 200 variable genes",
    color = "Cell type"
  )
```

## Try more variable genes

```{r}
variable_genes_2000 <- VariableFeatures(pbmc)[1:2000]
expr_2000 <- expr_t[, variable_genes_2000]
expr_2000_scaled <- scale(expr_2000, center = TRUE, scale = TRUE)

pca_2000 <- prcomp(expr_2000_scaled, center = FALSE, scale. = FALSE)
pca_2000_df <- as_tibble(pca_2000$x[, 1:2], rownames = "cell") %>%
  left_join(
    pbmc@meta.data %>% rownames_to_column("cell"),
    by = "cell"
  )
```

```{r}
ggplot(pca_2000_df, aes(x = PC1, y = PC2, color = seurat_annotations)) +
  geom_point(size = 0.7, alpha = 0.7) +
  labs(
    title = "PCA using top 2000 variable genes",
    color = "Cell type"
  )
```

Instructor pause:

Ask:

- Did the plot change?
- Why would gene selection affect PCA?
- Which version would you trust more, and why?

## UMAP with different number of PCs

```{r}
umap_5pcs <- umap(pca_scaled$x[, 1:5])
umap_50pcs <- umap(pca_scaled$x[, 1:50])
```

```{r}
umap_5_df <- as_tibble(umap_5pcs, .name_repair = "unique") %>%
  rename(UMAP1 = ...1, UMAP2 = ...2) %>%
  mutate(cell = rownames(pca_scaled$x)) %>%
  left_join(
    pbmc@meta.data %>% rownames_to_column("cell"),
    by = "cell"
  )

umap_50_df <- as_tibble(umap_50pcs, .name_repair = "unique") %>%
  rename(UMAP1 = ...1, UMAP2 = ...2) %>%
  mutate(cell = rownames(pca_scaled$x)) %>%
  left_join(
    pbmc@meta.data %>% rownames_to_column("cell"),
    by = "cell"
  )
```

```{r}
p_umap_5 <- ggplot(umap_5_df, aes(x = UMAP1, y = UMAP2, color = seurat_annotations)) +
  geom_point(size = 0.5, alpha = 0.7) +
  labs(title = "UMAP using 5 PCs") +
  theme(legend.position = "none")

p_umap_20 <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = seurat_annotations)) +
  geom_point(size = 0.5, alpha = 0.7) +
  labs(title = "UMAP using 20 PCs") +
  theme(legend.position = "none")

p_umap_50 <- ggplot(umap_50_df, aes(x = UMAP1, y = UMAP2, color = seurat_annotations)) +
  geom_point(size = 0.5, alpha = 0.7) +
  labs(title = "UMAP using 50 PCs")

p_umap_5 + p_umap_20 + p_umap_50
```

Instructor pause:

Ask:

- How much did the UMAP change?
- What does this tell us about parameter choices?
- Why should we report preprocessing choices?

# Module 6: Interpreting plots in practice

## Goal

By the end of this module, participants should be able to critically evaluate dimensionality reduction plots.

## Hide labels first

```{r}
ggplot(umap_df, aes(x = UMAP1, y = UMAP2)) +
  geom_point(size = 0.7, alpha = 0.7) +
  labs(title = "UMAP without labels")
```

Instructor activity:

Ask participants:

- How many groups do you see?
- Are all groups equally separated?
- Do any cells appear intermediate?
- What would you want to know before interpreting this?

## Reveal labels

```{r}
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = seurat_annotations)) +
  geom_point(size = 0.7, alpha = 0.7) +
  labs(
    title = "UMAP with cell-type annotations",
    color = "Cell type"
  )
```

Instructor pause:

Ask:

- Did the labels match expectations?
- Are any cell types mixed?
- Does overlap mean the labels are wrong?
- Does separation mean the biology is simple?

## Color by technical metrics

```{r}
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = nFeature_RNA)) +
  geom_point(size = 0.7, alpha = 0.7) +
  labs(
    title = "UMAP colored by number of detected genes",
    color = "nFeature_RNA"
  )
```

```{r}
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = percent.mt)) +
  geom_point(size = 0.7, alpha = 0.7) +
  labs(
    title = "UMAP colored by mitochondrial percentage",
    color = "percent.mt"
  )
```

Instructor note:

This is the “be careful” moment.

Emphasize:

> A plot can reflect biology, technical variation, or both.

# Final synthesis

## Core questions for interpreting any dimensionality reduction plot

Ask:

1. What are the observations?
2. What are the features?
3. What preprocessing was done?
4. What distance or similarity concept is implied?
5. What method was used?
6. What does the method preserve?
7. What labels or metadata were added after the fact?
8. Could technical factors explain the structure?
9. What follow-up analysis would you need?

# Optional hands-on exercise

Instructor note:

Use this only if there is time. Otherwise assign as homework.

## Student task

Use the same PBMC3k dataset and answer:

1. What happens if you use 500 variable genes instead of 1000?
2. What happens if you run UMAP using 10 PCs instead of 20?
3. Which plot gives the clearest cell-type separation?
4. Which plot would you be most cautious about interpreting?
5. What preprocessing choice had the biggest visible effect?

## Starter code

```{r eval=FALSE}
# Choose number of variable genes
n_genes <- 500

genes_use <- VariableFeatures(pbmc)[1:n_genes]

expr_use <- expr_t[, genes_use]
expr_use_scaled <- scale(expr_use, center = TRUE, scale = TRUE)

pca_use <- prcomp(expr_use_scaled, center = FALSE, scale. = FALSE)

pca_use_df <- as_tibble(pca_use$x[, 1:2], rownames = "cell") %>%
  left_join(
    pbmc@meta.data %>% rownames_to_column("cell"),
    by = "cell"
  )

ggplot(pca_use_df, aes(x = PC1, y = PC2, color = seurat_annotations)) +
  geom_point(size = 0.7, alpha = 0.7)
```

```{r eval=FALSE}
# Choose number of PCs
n_pcs <- 10

umap_use <- umap(pca_use$x[, 1:n_pcs])

umap_use_df <- as_tibble(umap_use, .name_repair = "unique") %>%
  rename(UMAP1 = ...1, UMAP2 = ...2) %>%
  mutate(cell = rownames(pca_use$x)) %>%
  left_join(
    pbmc@meta.data %>% rownames_to_column("cell"),
    by = "cell"
  )

ggplot(umap_use_df, aes(x = UMAP1, y = UMAP2, color = seurat_annotations)) +
  geom_point(size = 0.7, alpha = 0.7)
```

# Closing

Main takeaway:

> Dimensionality reduction is not just about making pretty plots.  
> It is about deciding how to represent, transform, summarize, and interpret high-dimensional biological data.
